---
title: "Multinomial Regression"
author: "Tate Huffman"
date: "4/27/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE)
library(magrittr)
library(gt)
library(haven)
library(tidyverse)
library(tidymodels)
```

## Multinomial Regression

Working with random forests was extremely computationally expensive, so here we're altering the model to see if it's feasible to just do multinomial logistic regression instead. The difference between multinomial regression and ordinary logistic regression is that the latter predicts only a binary outcome, while multinomial regression can be extended to predict more than just two outcomes.

```{r load_data}

# copying and altering most of code from explore.Rmd

load_files <- function(path) { 
  files <- dir(path, pattern = "data_\\d{4}.csv", full.names = TRUE)
  bind_rows(map_df(files, read_csv))
}

pitches <- load_files("data")

# creates pitch sequence

pitches %<>% mutate(pitch_seq = ifelse(!(pitch_number == 1 |
                                        is.na(pitch_type) |
                                        is.na(lag(pitch_type))), 
                                      paste0(lag(pitch_type), pitch_type),
                                      NA))

# adds in batter names, manually inputting unknown players

players <- read_csv("players.csv", col_types = cols()) %>% 
  select(mlb_id, mlb_name)

pitches %<>% 
  left_join(players, by = c("batter" = "mlb_id")) %>% 
  rename(batter_name = "mlb_name", pitcher_name = "player_name") %>% 
  mutate(batter_name = case_when(batter == 121347 ~ "Alex Rodriguez",
                                 batter == 116338 ~ "Torii Hunter",
                                 batter == 133380 ~ "Aramis Ramirez",
                                 batter == 120074 ~ "David Ortiz",
                                 batter == 150229 ~ "A.J. Pierzynski",
                                 batter == 218596 ~ "Tim Hudson",
                                 batter == 150359 ~ "A.J. Burnett",
                                 batter == 150302 ~ "Jason Marquis",
                                 batter == 329092 ~ "Randy Choate",
                                 batter == 279824 ~ "Mark Buehrle",
                                 batter == 605228 ~ "Jose Fernandez",
                                 batter == 115629 ~ "LaTroy Hawkins",
                                 TRUE ~ as.character(batter_name)))

# adding more pitch info in comparison to previous pitches

pitches %<>% 
  mutate(speed_diff = ifelse(!(pitch_number == 1 | 
                                 is.na(pitch_type) | 
                                 is.na(lag(pitch_type))),
                             release_speed - lag(release_speed), NA),
         spin_rate_diff = ifelse(!(pitch_number == 1 | 
                                 is.na(pitch_type) | 
                                 is.na(lag(pitch_type))),
                             release_spin_rate - lag(release_spin_rate), NA),
         loc_diff_x = ifelse(!(pitch_number == 1 | 
                                 is.na(pitch_type) | 
                                 is.na(lag(pitch_type))),
                             plate_x - lag(plate_x), NA),
         loc_diff_z = ifelse(!(pitch_number == 1 | 
                                 is.na(pitch_type) | 
                                 is.na(lag(pitch_type))),
                             plate_z - lag(plate_z), NA),
         loc_diff_total = ifelse(!(pitch_number == 1 | 
                                     is.na(pitch_type) | 
                                     is.na(lag(pitch_type))),
                                 sqrt(loc_diff_x^2 + loc_diff_z^2), NA))

# create binary indicator for whether pitcher was the starter
# also creates variable for score difference
# makes top/bottom of an inning a binary variable
# does the same for runners on base
# we also create an indicator for whether the pitcher/batter are same-handed
# we also turn the pitch sequences into factors here using fct_lump()
# does the same thing with pitches themselves

pitches %<>% 
  group_by(game_pk, inning_topbot) %>% 
  mutate(starter = ifelse(pitcher_name == first(pitcher_name), 1, 0)) %>% 
  ungroup() %>% 
  mutate(score_diff = fld_score - bat_score,
         inning_topbot = ifelse(inning_topbot == "Top", 1, 0),
         on_3b = ifelse(is.na(on_3b), 0, 1),
         on_2b = ifelse(is.na(on_2b), 0, 1),
         on_1b = ifelse(is.na(on_1b), 0, 1),
         same_handed = ifelse(stand == p_throws, 1, 0),
         stand = as_factor(stand),
         p_throws = as_factor(p_throws),
         pitch_seq = as_factor(pitch_seq), 
         pitch_seq = fct_lump(pitch_seq, prop = 0.02),
         pitch_type = as_factor(pitch_type), 
         pitch_type = fct_lump(pitch_type, prop = 0.02))

# we do more factor work with the description of different events
# grouping into ball, called strike, swinging strike, foul, and in play
# we create vectors categorizing each of these
# and then we mutate our tibble of pitches

ball <- c("ball", "blocked_ball", "intent_ball", "hit_by_pitch", "pitchout")
strike <- c("called_strike", "swinging_strike", "swinging_strike_blocked", 
            "missed_bunt", "swinging_pitchout", "foul", "foul_bunt", 
            "foul_tip", "foul_pitchout", "bunt_foul_tip")
in_play <- c("hit_into_play", "hit_into_play_no_out", "hit_into_play_score",
             "pitchout_hit_into_play_score")

pitches %<>% 
  mutate(description = case_when(description %in% ball ~ "ball",
                                 description %in% strike ~ "strike", 
                                 description %in% in_play ~ "in_play",
                                 TRUE ~ as.character(description)),
         description = as_factor(description))

# creates a smaller sample dataset w/ data from random 500 games

set.seed(1005)

pitch_sample <- pitches %>% 
  filter(game_pk %in% c(sample(pull(select(pitches, game_pk)), 300)))

# creating new tibbles with this data
# focusing on sequences and outcomes means we focus only on the end result of
# PAs featuring at least 2 pitches
# so this analysis definitely has holes, given that it's ignoring lots of data
# but as a rough, preliminary analysis, hopefully it holds up

pitches_mod <- pitches %>%
  filter(!is.na(pitch_seq), !is.na(release_spin_rate))

pitch_sample_mod <- pitch_sample %>% 
  filter(!is.na(pitch_seq), !is.na(release_spin_rate))

```

The goal is to accurately predict whether a pitch will be a ball, strike, or hit into play based on characteristics given. We'll create three models and compare them: the first a simple model generating predictions solely from pitch sequence and its associated characteristics (adjusting for batter/pitcher handedness), the second incorporating more granular information about the pitch itself, and the third adjusting for game context. 

```{r model_prep}

# a fair chunk of this code comes from PPBDS
# this first model sees how well we can predict outcome based on comparisons
# with the previous pitch

seq_form <- formula(description ~ pitch_seq * same_handed + speed_diff + 
                      loc_diff_x + loc_diff_z + spin_rate_diff)

# this model includes more information about the pitch itself (not relative)

pitch_form <- update(seq_form, ~ . + release_speed + release_spin_rate + 
                       pfx_x + pfx_z + plate_x + plate_z)

# this model includes contextual information
# specifically, count, outs, inning, baserunners, and score differential
# as well as whether the pitcher is a starter

context_form <- update(pitch_form, ~ . + balls + strikes + outs_when_up +
                         on_3b + on_2b + on_1b + inning + inning_topbot +
                         starter)

# here, we combine all these models into a single tibble

pitch_formulas <- tibble(formula = c(seq_form,
                                     pitch_form,
                                     context_form),
                         group = c("Sequence model",
                                   "Pitch type model",
                                   "Context model"))

# create specifications for classification
# we also create the random forest specification, just in case

multinom_spec <- multinom_reg() %>% 
  set_engine("glmnet") %>% 
  set_mode("classification")

rf_spec <- rand_forest() %>%
  set_engine("randomForest") %>%
  set_mode("classification") 

```

Now that we've set ourselves up with this model, let's start our process of analysis. First, we'll split the data into train and test sets.

```{r train_test}

# split model into train and test sets
# first cleans data to exclude NA values

pitch_sample_mod %<>% 
  filter(!is.na(release_speed & pfx_x & pfx_z & plate_x & plate_z &
                  speed_diff & loc_diff_x & loc_diff_z))

set.seed(1005)

pitch_sample_split <- initial_split(pitch_sample_mod, prop = 0.8)
pitch_sample_train <- training(pitch_sample_split)
pitch_sample_test <- testing(pitch_sample_split)

```
